---
title: "ggplot2 Blog Post"
author: "Granger Moch"
date: "September 29, 2014"
output: pdf_document
fig_caption: yes
fig_height: 8
fig_width: 9
number_sections: yes
toc: yes
abstract: Demonstrate the functionality of ggplot and proide an introduction to its syntax and uses.
---

\newpage 

#Purpose

I am creating this guide not solely for the assignment, but so that those that are new to R can recognize the extensive functionality it offers. Furthermore, I hope to provide a succinct introduction to several different graph types and their uses. for both my sake as well as those who read this guide.

#Data

## Simulate 3 Factor Variables
```{r}
FacVar1=as.factor(rep(c("level1","level2"),25))
FacVar2=as.factor(rep(c("levelA","levelB","levelC"),17)[-51])
FacVar3=as.factor(rep(c("levelI","levelII","levelIII","levelIV"),13)[-c(51:52)])
```

## Simulate 4 Numeric Variables
```{r}
set.seed(123)
NumVar1=round(rnorm(n=50,mean=1000,sd=50),digits=2) ## Normal distribution
set.seed(123)
NumVar2=round(runif(n=50,min=500,max=1500),digits=2) ## Uniform distribution
set.seed(123)
NumVar3=round(rexp(n=50,rate=.001)) ## Exponential distribution
NumVar4=2001:2050

simData=data.frame(FacVar1,FacVar2,FacVar3,NumVar1,NumVar2,NumVar3,NumVar4)
```
##The ggplot2 and reshape2 libraries need to be initialized for use on this page

```{r}
library(ggplot2)
library(reshape2)
```
##General Syntax

Each ggplot2 plot will begin with the function ggplot(), which has two primary arguments:


- data
The data frame containing the data to be plotted  

 
- aes()
The aesthetic mappings to assign to plot elements

The **geom_()** functions are used alongside the  + operator, in order to assign a geometric object to represent the data

**Syntax:**

>```ggplot(dataset, aes(x, y))+geom_point()```

\newpage 

#One Variable: Numeric Variable

##Index Plot of One Numeric Variable
```{r}
ggplot(simData,aes(y=NumVar1,x=1:nrow(simData),group="NumVar1"))+geom_point()+geom_line()+ xlab("") ## Index plot 
```
   
   
     
-Axis labels can be changed using **xlab()** and **ylab()** respectively for the x and y axes. In this instance, the x axis label is blank as indicated by the null value.


\newpage

##Histogram of One Numeric Variable
```{r}
ggplot(simData,aes(x=NumVar1))+geom_histogram()
```

-This plot uses the geom_histogram(), for a list of all geoms available for use see here: http://sape.inf.usi.ch/quick-reference/ggplot2/geom

\newpage

##Kernel Density Plot of One Numeric Variable
```{r}
ggplot(simData,aes(x=NumVar1))+geom_density() 
```
\newpage

##Box Plot of One Numeric Variable
```{r}
ggplot(simData,aes(x=factor(""),y=NumVar1))+geom_boxplot()+ xlab("") ## box plot
```
\newpage

#Plotting One Variable: Factor Variable

## Barplot of One Factor Variable
```{r}
ggplot(simData,aes(x=FacVar3))+geom_bar()
```
\newpage

## Pie chart of One Factor Variable
```{r}
ggplot(simData,aes(x = factor(""), fill=FacVar3, label=FacVar3))+geom_bar()+ coord_polar(theta = "y")  +scale_x_discrete("")
```

The coordinate polar system (coord_polar) is most typically used for pie charts, as seen above.

**Syntax:**

coord_polar(theta = "x", start = 0, direction = 1)

**Arguments:**
-*theta*= variable to map angle to (x or y) 
-*start*= offset of starting point from 12 o'clock in radians
-*direction*= 1, clockwise; -1, anticlockwise

-In this instance, the variable angle is mapped to y.
\newpage 

#Two Variables: Two Numeric Variables

##Line Plot of Two Numeric Variables
```{r}
simtmp=simData[,c(4:5)] ## 4th and 5th columns are NumVar1 and NumVar2
simtmp$index=1:nrow(simtmp)
simtmpmelt=melt(simtmp,id=c("index"))

## line plots with observation number as index
ggplot(simtmpmelt,aes(y=value,x=index,color=variable))+geom_point()+geom_line()+xlab("")
```

 

With this lineplot, the lines are colored according to the variable.  This is specified as an arugment in the aes() function as "color=variable." 

The **melt** function, as part of the Reshape2 package, essentially "melts" the data you provide it so that each row is a unique id-variable combination. In other words, the "wide-format" data is "melted" into what is called "long-form data". As these formats relate to ggplot, ggplot2 actually requires long-form data(also known as *tidy data*); this is why the **melt** function was used here. The other core function of Reshape2 is **cast**, which does just the opposite. **Cast** takes long-format data and casts it into wide-format data.
\newpage 

##Density Plot of Two Numeric Variables

```{r}
ggplot(simtmpmelt,aes(x=value,color=variable))+geom_density()
```

This Density Plot uses the simptmpmelt dataset created on the previous page.

\newpage 

##Scatter Plot of Two Numeric Variables
```{r}
ggplot(simData,aes(x=NumVar1,y=NumVar2))+geom_point()
```

\newpage 

#Two Variables: Two Factor Variables

##Mosaic Plot of Two Factor Variables
```{r}
ggMMplot <- function(var1, var2){
  require(ggplot2)
  levVar1 <- length(levels(var1))
  levVar2 <- length(levels(var2))
  
  jointTable <- prop.table(table(var1, var2))
  plotData <- as.data.frame(jointTable)
  plotData$marginVar1 <- prop.table(table(var1)) ##creates a table of proportions
  plotData$var2Height <- plotData$Freq / plotData$marginVar1
  plotData$var1Center <- c(0, cumsum(plotData$marginVar1)[1:levVar1 -1]) +
    plotData$marginVar1 / 2
  
  ggplot(plotData, aes(var1Center, var2Height)) +
    geom_bar(stat = "identity", aes(width = marginVar1, fill = var2), col = "Black") +
    geom_text(aes(label = as.character(var1), x = var1Center, y = 1.05)) 
}
ggMMplot(simData$FacVar2, simData$FacVar3)
```

**Note:Because ggplot2 does not currently support mosaic plots, the function ggMMplot is created from scratch to create the plot.**

\newpage 

##Bar Plot of Two Factor Variables
```{r}
bartabledat = as.data.frame(table(simData$FacVar2, simData$FacVar3)) ## get the cross tab
ggplot(bartabledat,aes(x=Var2,y=Freq,fill=Var1))+geom_bar(position="dodge", stat="identity") 
## added stat="identity" (since we are in fact, mapping a variable to y)
```
\newpage 

##Stacked Bar Plot of Two Factor Variables
```{r}
ggplot(bartabledat,aes(x=Var2,y=Freq,fill=Var1))+geom_bar(stat="identity") ## stacked
```

\newpage 

##Stacked Bar Plot of Two Factor Variables (100%)
```{r}
bartableprop =as.data.frame(prop.table(table(simData$FacVar2, simData$FacVar3),2)*100) 
ggplot(bartableprop,aes(x=Var2,y=Freq,fill=Var1))+geom_bar(stat="identity")##added stat="identity"
```
\newpage 

#Two Variables: One Factor and One Numeric Variable

##Box Plot of One Factor and One Numeric Variable
```{r}
ggplot(simData,aes(x=FacVar1,y=NumVar1))+geom_boxplot()
```
\newpage 

##Density Plot of One Factor and One Numeric Variable
```{r}
ggplot(simData,aes(x=NumVar1,color=FacVar1))+geom_density()
```
\newpage 

##Dot Chart of One Factor and One Numeric Variable
```{r}
meanagg = aggregate(simData$NumVar1, list(simData$FacVar3), mean) ##aggregate function groups data by the specified variables
ggplot(meanagg,aes(x=Group.1,y=x))+geom_point()+coord_flip() ## Dot Chart equivalent
``` 
 
The **coord_flip()** function flips cartesian coordinates so that horizontal becomes vertical and vice-versa. This is particularly useful for creating boxplots and other horizontal, interval geoms.
\newpage 

##Bar Plot of One Factor and One Numeric Variable
```{r}
meanagg = aggregate(simData$NumVar1, list(simData$FacVar3), mean)
ggplot(meanagg,aes(x=Group.1,y=x))+geom_bar(stat="identity")
##added stat="identity"
```
\newpage 

#Three Variables: Three Factor Variables

##Bar Plot of Three Factor Variables
```{r}
Threebartable = as.data.frame(table(simData$FacVar1, simData$FacVar2, simData$FacVar3)) ## CrossTab
ggplot(Threebartable,aes(x=Var3,y=Freq,fill=Var2))+geom_bar(position="dodge",stat="identity")+facet_wrap(~Var1) ## Bar plot with facetting
```

**The facet_wrap() function wraps a 1d ribbon of panels into a 2d ribbon.**


\newpage 

#Three Variables: One Numeric and Two Factor Variables
##Box Plot of One Numeric and Two Factor Variables
```{r}
## boxplot of NumVar1 over an interaction of 6 levels of the combination of FacVar1 and FacVar2
ggplot(simData,aes(x=FacVar2,y=NumVar1, fill=FacVar1))+geom_boxplot()
```
\newpage 

##Dotchart of One Numeric and Two Factor Variables
```{r}
## Mean of 1 Numeric over levels of two factor vars
meanaggg = aggregate(simData$NumVar1, list(simData$FacVar1, simData$FacVar2), mean)
## Dot Chart equivalent
ggplot(meanaggg,aes(x=Group.2,y=x,color=Group.2))+geom_point()+coord_flip()+facet_wrap(~Group.1, ncol=1)
```
\newpage 

##Interaction Line Chart of One Numeric and Two Factor Variables
```{r}
## Interaction chart - line chart
ggplot(meanaggg,aes(x=Group.2,y=x,color=Group.1, group=Group.1))+geom_point()+geom_line()
```
\newpage

##Bar Plot of One Numeric and Two Factor Variables 
```{r}
ggplot(meanaggg,aes(x=Group.2,y=x))+geom_bar(stat="identity")+facet_wrap(~Group.1)
```
\newpage

#Three Variables: Two Numeric and One Factor Variable
##Scatter Plot of Two Numeric and One Factor Variable
```{r}
ggplot(simData,aes(x=NumVar1,y=NumVar2,color=FacVar1))+geom_point()
## Scatter plot with color identifying the factor variable
```

\newpage
#Three Variables: Three Numeric Variables
##Name this Graph
```{r}
# NumVar4 is 2001 through 2050... possibly, a time variable - use that as the x-axis
simtmpp=simData[,c(4,5,7)]
simtmppmelt=melt(simtmpp,id=c("NumVar4"))
ggplot(simtmppmelt,aes(x=NumVar4,y=value,color=variable,group=variable))+geom_point()+geom_line()
```
\newpage

##Stacked Area Graph of Three Numeric Variables
```{r}
ggplot(simtmppmelt,aes(x=NumVar4,y=value,fill=variable))+geom_area(position="stack")
```

\newpage

##Stacked Area Graph of Three Numeric Variables(100%)
```{r}
ggplot(simtmppmelt,aes(x=NumVar4,y=value,fill=variable))+geom_area(position="fill")
```



